Beyond ``Horizontal'' and ``Vertical'':\\ The Welfare Effects of Complex Integration\tnoteref{t1}

### 2023-06-02 doc/VerticalArrangements.tex
revised draft that addresses key comments from IJIO reviewers. Converted format of doc back to basic latex from elsevier

\documentclass[12pt,review,authoryear]{elsarticle}
%\usepackage{ecrc}       % needed for elsarticle
%\usepackage{chicago}    % bibliography package
\usepackage{graphicx}   % insert PostScript figures
\usepackage{setspace}   % controls line spacing
\usepackage{amsmath,amsthm,amssymb,amstext} % controls equation entry and symbols
\usepackage{rotating}   % rotates graphics
\usepackage{soul}       % controls hyphenation
\usepackage{epsfig}     % helps with including graphics
\usepackage{pdflscape}  % helps with displaying rotated graphics in PDFs
\usepackage{lscape}     % helps with rotating pages
\usepackage{caption}    % controls  captions
\usepackage{adjustbox}  % shrink tables
\usepackage[margin=1in]{geometry} % margins
\usepackage{hyperref}   % displays URLs
%packages for kable tables
 \usepackage{booktabs}
 \usepackage{longtable}
 \usepackage{array}
 \usepackage{multirow}
 \usepackage{wrapfig}
 \usepackage{float}
 \usepackage{colortbl}
 \usepackage{pdflscape}
 \usepackage{tabu}
 \usepackage{threeparttable}
 \usepackage{threeparttablex}
 \usepackage[normalem]{ulem}
 \usepackage{makecell}
 \usepackage{xcolor}

@@

@@

\begin{frontmatter}

%% Title, authors and addresses

%% use the tnoteref command within \title for footnotes;
%% use the tnotetext command for the associated footnote;
%% use the fnref command within \author or \address for footnotes;
%% use the fntext command for the associated footnote;
%% use the corref command within \author for corresponding author footnotes;
%% use the cortext command for the associated footnote;
%% use the ead command for the email address,
%% and the form \ead[url] for the home page:
%%
%% \title{Title\tnoteref{label1}}
%% \tnotetext[label1]{}
%% \author{Name\corref{cor1}\fnref{label2}}
%% \ead{email address}
%% \ead[url]{home page}
%% \fntext[label2]{}
%% \cortext[cor1]{}
%% \address{Address\fnref{label3}}
%% \fntext[label3]{}

%%\dochead{}
%% Use \dochead if there is an article header, e.g. \dochead{Short communication}

\title{Beyond ``Horizontal'' and ``Vertical'':\\ The Welfare Effects of Complex Integration\tnoteref{t1} }
\tnotetext[t1]{The analysis and conclusions set forth are those of the authors and do not indicate concurrence by other members of the Board research staff or by the Federal Reserve Board of Governors. Furthermore, the views expressed here should not be purported to reflect those of the U.S. Department of Justice. This article has benefited from conversations with Alison Oldale, Ted Rosenbaum, and Nathan Wilson.}

%% use optional labels to link authors explicitly to addresses:
%% \author[label1,label2]{<author name>}
%% \address[label1]{<address>}
%% \address[label2]{<address>}

\author[1]{Margaret Loudermilk\fnref{fn1}}
\ead{margaret.loudermilk@usdoj.gov}
\affiliation[1]{organization={US Department of Justice}, addressline= {450 5th St. NW}, city={Washington, DC},postcode={20530}}
\author[2]{Gloria Sheu\fnref{fn1}}
\ead{gloria.sheu@frb.gov}
\affiliation[2]{organization={Federal Reserve Board}, addressline= { 2051 Constitution Ave. NW}, city={Washington, DC},postcode={20418} }
\author[3]{Charles Taragin\corref{cor1}\fnref{fn1}}
 \cortext[cor1]{Corresponding author.}
\ead{charles.s.taragin@frb.gov}
\affiliation[3]{organization={Federal Reserve Board}, addressline= { 2051 Constitution Ave. NW}, city={Washington, DC},postcode={20418} }
\fntext[fn1]{Declarations of interest: none.}

@@

\begin{keyword}
bargaining models\sep merger simulation\sep vertical markets\sep vertical mergers
\JEL  L13\sep L40\sep L41\sep L42
\end{keyword}

\end{frontmatter}

%\pagenumbering{roman}       % Roman numerals from abstract to text
%\maketitle                  % print title information
%\thispagestyle{empty}       % no page number on THIS page

%\newpage                    % start a new page
%\pagenumbering{arabic}      % Arabic page numbers from now on
%\doublespacing

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Body of Paper
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
Given the many different ways a firm may be organized, it is rare that any particular merger can be neatly categorized as purely ``horizontal'' or ``vertical.''  This is particularly true for the mergers involving large firms that usually attract the most scrutiny.  However, for both historical and practical reasons, antitrust research and practice has generally maintained this dichotomy.  Horizontal mergers are cast in terms of whether they create significant upward pricing pressure (UPP), whereas the focus for vertical mergers is on assessing the net effects of the elimination of double marginalization (EDM) versus raising rivals' costs (RRC).  As a result, the antitrust community has little consensus for what welfare impacts one should expect from a merger that combines all of these effects simultaneously.

The horizontal versus vertical distinction has been enshrined in U.S. Department of Justice (DOJ) and Federal Trade Commission (FTC) Horizontal Merger Guidelines and their companion document, the Vertical Merger Guidelines.\footnote{The FTC withdrew its approval from the Vertical Merger Guidelines in 2021.  The DOJ did not.  }  Given these policy documents, it is not surprising that the agencies typically frame most mergers as either horizontal or vertical.  However, as part of the 2022 effort by the DOJ and the FTC to revise their merger guidelines, these agencies requested comments on whether any ``modern market realities may be lost by focusing on [horizontal and vertical] relationships categorically.''\footnote{The full question is 1.g. in the ``Request for Information on Merger Enforcement,'' dated January 18, 2022.  The text of the question reads: ``Should the guidelines' traditional distinctions between horizontal and vertical mergers be revisited in light of recent economic trends in the modern economy? What aspects of modern market realities may be lost by focusing on these relationships categorically? Should the guidelines address all mergers in a common framework that covers all market relationships relevant to competition? If so, how?''.  }  In this paper, our goal is to provide guidance to practitioners on how these mergers can be simulated in a familiar framework and to also show the range of welfare outcomes that are likely given this modeling environment.

Our model is drawn directly from \cite{ST2021}, and features a Bertrand logit downstream setup alongside a Nash Bargaining wholesale negotiation upstream.  We use this model to numerically simulate four types of mergers: (1) vertical mergers between an unintegrated retailer and an unintegrated wholesaler, (2) downstream ``horizontal'' mergers between an unintegrated retailer and an integrated retailer/wholesaler pair, (3) upstream ``horizontal'' mergers between an unintegrated wholesaler and an integrated retailer/wholesaler pair, and (4) integrated mergers between two previously integrated retailer/wholesaler pairs.  In each of these simulations, we calculate the impacts on consumer welfare.  Furthermore, we examine how the presence of additional integrated rivals (separate from the merging firms) changes our results.

Our simulations show certain patterns across mergers.  For horizontal upstream and downstream mergers, we find that consumers are net harmed in most instances.  The exceptions occur when upstream wholesalers have more bargaining power than downstream retailers, which is when we expect pre-merger input prices to be high and the gains from EDM to be larger.  Furthermore, we find that integrated mergers are typically net harmful to consumers as well.  The exceptions happen only when upstream firms have significantly more bargaining power compared to downstream firms.  Thus, although these mergers have both horizontal and vertical aspects, they appear to be qualitatively similar to pure horizontal mergers in their overall consumer impacts.  Furthermore, we find that the presence of integrated rival firms lowers the spread of welfare outcomes for consumers and slightly shifts these outcomes towards less consumer harm.  The existence of rivals with their own sources of bargaining leverage stemming from vertical integration appears to reduce the variance of merger outcomes, but does not substantially mitigate consumer harm.

We then employ this model to investigate the 2021 merger of two solid waste management firms, Republic Services and Santek Waste Services, in one geographic market, Chattanooga, Tennessee. The supply chain of this industry involves collection haulers at the downstream level that bring trash to disposal facilities at the upstream level.  Both Republic and Santek operated upstream and downstream in Chattanooga, as did one of their largest rivals, Waste Connections.  In this market, the DOJ negotiated a settlement package meant to address harms in downstream collection and in upstream disposal.  Using a small set of public data inputs, our merger simulation model is able to produce rich results for the changes in horizontal and vertical competition, including RRC and EDM effects.  We estimate that without divestitures, consumers would have faced approximately \$16 million in annual harm in the Chattanooga market.  Furthermore, we show that modeling the merger as a purely horizontal upstream or downstream transaction would miss key vertical impacts.

As a caveat to our analysis, it is based on one specific model of competition and mergers.  Certain alternative structures are likely to give different welfare results.  Indeed, we found in \cite{ST2021} that using a second score auction to model downstream competition, as opposed to assuming Bertrand pricing, could meaningfully impact the estimated consumer harm from mergers, although results were similar between these models for vertical mergers.  This paper is an attempt to provide some information about possible welfare effects from complex mergers within a framework that is used by practitioners.  Exploring other commonly used frameworks would be a natural extension in future research.

Our paper is related to the literature on methods for measuring the welfare impacts of mergers, particularly papers about vertical mergers.  The closest merger simulation article is our previous work in \cite{ST2021}, which uses the same model we feature here.  That paper in turn takes the setup from \cite{WF1994} to characterize downstream competition, and pairs it with a Nash Bargaining model for the upstream market.  \cite{DVDS2018} and \cite{DS2019} provide additional vertical merger frameworks that make alternative assumptions about demand and pricing.  \cite{MS2013} and \cite{Rogerson2020} develop UPP-style tools for assessing the impact of vertical mergers.  However, none of these papers have results for the wide range of merger types that we study.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Theory\label{sec:theory}}
We begin by describing our basic framework, taken from \cite{ST2021}, which uses a downstream Bertrand logit model embedded in an upstream Nash Bargaining setup.  This framework allows us to study a variety of merger configurations.

@@

Together the series of downstream and upstream first order conditions determine market equilibrium.  This model can be solved and calibrated as described in \cite{ST2021}.  In our baseline configuration, we assume that wholesale and retail marginal costs are constant.  We allow for linear increasing marginal costs as a robustness check that appears in Appendix \ref{app:mc}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Mergers}
Here we describe the manner in which mergers are modeled in this framework.  We begin by describing vertical mergers, as all cases we examine have some vertical aspects.  Suppose that retailer $r$ and wholesaler $w$ were to merge.  Then the first order condition for profit maximization for the joint firm when setting the downstream price for product $rw$ is given by
\begin{equation}
\begin{split}
&\sum_{x \in \mathbb{W}^r \setminus \{w\}} [p_{rx} - p^W_{rx} - c^R_{rx}]\frac{\partial s_{rx}}{\partial p_{rw}} + s_{rw} + \overbrace{[p_{rw} - c^W_{rw} - c^R_{rw}]\frac{\partial s_{rw}}{\partial p_{rw}}}^{\text{EDM effect}} \\
&+ \underbrace{\sum_{t \in \mathbb{R}^w \setminus \{r\}} [p^W_{tw} - c^W_{tw}] \frac{\partial s_{tw}}{\partial p_{rw}}}_{\text{upstream UPP effect}} = 0.
\end{split}
\label{eq: vmerger own ret downstream foc}
\end{equation}
The pricing problem balances two effects.  On the one hand, the term labeled ``EDM effect'' captures the impact of retailer $r$ being able to access product $w$ at marginal cost.  This force would tend to lower the resulting price.  On the other hand, the term labeled ``upstream UPP effect'' captures the incentive to raise prices for retailer $r$ in order to divert sales to wholesaler $w$.  This force would tend to raise the resulting consumer price.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Numerical Simulations \label{sec:sims}}
In this section, we use a similar simulation setup to that in \cite{ST2021}, but allow for additional types of mergers beyond purely horizontal and purely vertical combinations.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Data Generating Process}
Broadly speaking, we consider four categories of mergers: downstream, upstream, vertical, and integrated.  We define a downstream merger as a merger between two unintegrated retailers or between an unintegrated retailer and a vertically integrated wholesaler/retailer combination.  Similarly, upstream mergers are those between two unintegrated wholesalers or between an unintegrated wholesaler and a vertically integrated wholesaler/retailer.  We define vertical mergers as those between an unintegrated wholesaler and unintegrated retailer.  Finally, an integrated merger is between two firms that are both already vertically integrated pre-merger.

@@

This paper relaxes an assumption made in much of the merger simulation literature: that mergers occur in vertical supply chains where in the pre-merger state, no firms are vertically integrated. Removing this limitation is important as competition authorities are routinely called upon to investigate mergers where either the merging parties or a third party are vertically integrated. Using a series of numerical simulations and the model of \cite{ST2021}, we find that most mergers involving one or more firms that are already vertically integrated harm consumers.  The exceptions occur when wholesalers have more bargaining power compared to retailers, which are instances where the benefits from EDM would tend to be larger.  We also show that having integrated non-merging rivals limits the range of welfare outcomes and tends to make mergers somewhat more beneficial to consumers, although these firms' presence is not a panacea for consumer harm. We apply our model to the Republic/Santek merger, where we find that ignoring existing vertical relationships gives an incomplete picture of the merger's welfare impacts.

Our results suggest that mergers that have horizontal aspects remain harmful for consumers in many instances, even when these mergers also have vertical aspects.
This appears to be especially true when retailers have relatively more bargaining power, which in our model occurs when retailer margins are relatively large compared to wholesaler margins.  Antitrust agencies may have reason to be especially cautious when reviewing mergers in these situations.  However, we of course have the caveat that our results are specific to our modeling framework.  An important question for future research would be to explore merger outcomes in other contexts.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Appendix: Increasing Marginal Costs\label{app:mc}}

>>>>

\documentclass[12pt]{article}
%\documentclass[12pt,review,authoryear]{elsarticle}
%\usepackage{ecrc}       % needed for elsarticle
%\usepackage{chicago}    % bibliography package
\usepackage{graphicx}   % insert PostScript figures
\usepackage{setspace}   % controls line spacing
\usepackage{amsmath,amsthm,amssymb,amstext} % controls equation entry and symbols
\usepackage{rotating}   % rotates graphics
\usepackage{soul}       % controls hyphenation
\usepackage{epsfig}     % helps with including graphics
\usepackage{pdflscape}  % helps with displaying rotated graphics in PDFs
\usepackage{lscape}     % helps with rotating pages
\usepackage{caption}    % controls  captions
\usepackage{adjustbox}  % shrink tables
\usepackage[margin=1in]{geometry} % margins
\usepackage{hyperref}   % displays URLs
%packages for kable tables
 \usepackage{booktabs}
 \usepackage{longtable}
 \usepackage{array}
 \usepackage{multirow}
 \usepackage{wrapfig}
 \usepackage{float}
 \usepackage{colortbl}
 \usepackage{pdflscape}
 \usepackage{tabu}
 \usepackage{threeparttable}
 \usepackage{threeparttablex}
 \usepackage[normalem]{ulem}
 \usepackage{makecell}
 \usepackage{xcolor}

@@

% \begin{frontmatter}
%
% %% Title, authors and addresses
%
% %% use the tnoteref command within \title for footnotes;
% %% use the tnotetext command for the associated footnote;
% %% use the fnref command within \author or \address for footnotes;
% %% use the fntext command for the associated footnote;
% %% use the corref command within \author for corresponding author footnotes;
% %% use the cortext command for the associated footnote;
% %% use the ead command for the email address,
% %% and the form \ead[url] for the home page:
% %%
% %% \title{Title\tnoteref{label1}}
% %% \tnotetext[label1]{}
% %% \author{Name\corref{cor1}\fnref{label2}}
% %% \ead{email address}
% %% \ead[url]{home page}
% %% \fntext[label2]{}
% %% \cortext[cor1]{}
% %% \address{Address\fnref{label3}}
% %% \fntext[label3]{}
%
% %%\dochead{}
% %% Use \dochead if there is an article header, e.g. \dochead{Short communication}
%
% \title{Beyond ``Horizontal'' and ``Vertical'':\\ The Welfare Effects of Complex Integration\tnoteref{t1} }
% \tnotetext[t1]{The analysis and conclusions set forth are those of the authors and do not indicate concurrence by other members of the Board research staff or by the Federal Reserve Board of Governors. Furthermore, the views expressed here should not be purported to reflect those of the U.S. Department of Justice. This article has benefited from conversations with Alison Oldale, Ted Rosenbaum, and Nathan Wilson.}
%
% %% use optional labels to link authors explicitly to addresses:
% %% \author[label1,label2]{<author name>}
% %% \address[label1]{<address>}
% %% \address[label2]{<address>}
%
% \author[1]{Margaret Loudermilk\fnref{fn1}}
% \ead{margaret.loudermilk@usdoj.gov}
% \affiliation[1]{organization={US Department of Justice}, addressline= {450 5th St. NW}, city={Washington, DC},postcode={20530}}
% \author[2]{Gloria Sheu\fnref{fn1}}
% \ead{gloria.sheu@frb.gov}
% \affiliation[2]{organization={Federal Reserve Board}, addressline= { 2051 Constitution Ave. NW}, city={Washington, DC},postcode={20418} }
% \author[3]{Charles Taragin\corref{cor1}\fnref{fn1}}
%  \cortext[cor1]{Corresponding author.}
% \ead{charles.s.taragin@frb.gov}
% \affiliation[3]{organization={Federal Reserve Board}, addressline= { 2051 Constitution Ave. NW}, city={Washington, DC},postcode={20418} }
% \fntext[fn1]{Declarations of interest: none.}

\title{Beyond ``Horizontal'' and ``Vertical'':\\ The Welfare Effects of Complex Integration\footnote{The analysis and conclusions set forth are those of the authors and do not indicate concurrence by other members of the Board research staff or by the Federal Reserve Board of Governors. Furthermore, the views expressed here should not be purported to reflect those of the U.S. Department of Justice. This article has benefited from conversations with Alison Oldale, Ted Rosenbaum, and Nathan Wilson.}}

\newcommand*\samethanks[1][\value{footnote}]{\footnotemark[#1]}
\author{Margaret Loudermilk\footnote{margaret.loudermilk@usdoj.gov} \\  U.S Department of Justice \and Gloria Sheu\footnote{gloria.sheu@frb.gov. } \\ Federal Reserve Board \and  Charles Taragin\footnote{charles.s.taragin@frb.gov (corresponding author).}  \\  Federal Reserve Board }

@@

\pagenumbering{roman}       % Roman numerals from abstract to text
\maketitle                  % print title information
\thispagestyle{empty}       % no page number on THIS page

@@

% \begin{keyword}
% bargaining models\sep merger simulation\sep vertical markets\sep vertical mergers
% \JEL  L13\sep L40\sep L41\sep L42
% \end{keyword}
%
% \end{frontmatter}

\bigbreak Keywords: bargaining models; merger simulation; vertical markets; vertical mergers

JEL classification: L13; L40; L41; L42

\newpage                    % start a new page
\pagenumbering{arabic}      % Arabic page numbers from now on
\doublespacing

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Body of Paper
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
Given the many different ways a firm may be organized, it is rare that any particular merger can be neatly categorized as purely ``horizontal'' or ``vertical.''  This is particularly true for the mergers involving large firms that usually attract the most scrutiny.  However, for both historical and practical reasons, antitrust research and practice has generally maintained this dichotomy.  Horizontal mergers are cast in terms of whether they create significant upward pricing pressure (UPP), whereas the focus for vertical mergers is on assessing the net effects of the elimination of double marginalization (EDM) versus raising rivals' costs (RRC).  As a result, the antitrust community has little consensus regarding how complex mergers that possess both horizontal and vertical aspects will affect welfare.

The horizontal versus vertical distinction has been enshrined in U.S. Department of Justice (DOJ) and Federal Trade Commission (FTC) Horizontal Merger Guidelines and their companion document, the Vertical Merger Guidelines.\footnote{The FTC withdrew its approval from the Vertical Merger Guidelines in 2021.  The DOJ did not.  }  Given these policy documents, it is not surprising that the agencies typically frame most mergers as either horizontal or vertical.  However, as part of the 2022 effort by the DOJ and the FTC to revise their merger guidelines, these agencies requested comments on whether any ``modern market realities may be lost by focusing on [horizontal and vertical] relationships categorically.''\footnote{The full question is 1.g. in the ``Request for Information on Merger Enforcement,'' dated January 18, 2022.  The text of the question reads: ``Should the guidelines' traditional distinctions between horizontal and vertical mergers be revisited in light of recent economic trends in the modern economy? What aspects of modern market realities may be lost by focusing on these relationships categorically? Should the guidelines address all mergers in a common framework that covers all market relationships relevant to competition? If so, how?''.  }  In this paper, we address this question using a vertical supply chain model to explore the welfare implications of complex mergers.

Our model is drawn directly from \cite{ST2021}, and features a Bertrand logit downstream setup alongside a Nash Bargaining wholesale negotiation upstream.  We use this model to explore the welfare implications of four types of complex mergers: (1) vertical mergers between an unintegrated retailer and an unintegrated wholesaler, (2) downstream ``horizontal'' mergers between an unintegrated retailer and an integrated retailer/wholesaler pair, (3) upstream ``horizontal'' mergers between an unintegrated wholesaler and an integrated retailer/wholesaler pair, and (4) integrated mergers between two previously integrated retailer/wholesaler pairs.

Our model-based simulations show certain patterns across mergers.  For horizontal upstream and downstream mergers, we find that the consumer harm from UPP and RRC typically exceeds any gains from EDM.  The exceptions occur when upstream wholesalers have more bargaining power than downstream retailers, which is when we expect pre-merger input prices to be high and the gains from EDM to be larger.  Likewise, we find that integrated mergers are typically net harmful to consumers as well.  The exceptions happen only when upstream firms have significantly more bargaining power compared to downstream firms.  Thus, although these complex mergers have both horizontal and vertical aspects, they appear to be qualitatively similar to pure horizontal mergers in their overall consumer impacts.

We also examine how the presence of additional integrated rivals (separate from the merging firms) changes our results. We find that the presence of integrated rival firms lowers the spread of welfare outcomes for consumers and slightly shifts these outcomes towards less consumer harm.  The existence of rivals with their own sources of bargaining leverage stemming from vertical integration appears to reduce the variance of merger outcomes, but does not substantially mitigate consumer harm.

We then employ this model to investigate the 2021 merger of two vertically integrated solid waste management firms, Republic Services and Santek Waste Services, in one geographic market, Chattanooga, Tennessee. The supply chain of this industry involves collection haulers at the downstream level that bring trash to disposal facilities at the upstream level.  Both Republic and Santek operated upstream and downstream in Chattanooga, as did one of their largest rivals, Waste Connections.  In this market, the DOJ negotiated a settlement package meant to address harms in downstream collection and in upstream disposal.  Using a small set of public data inputs, our merger simulation model is able to produce rich results for the changes in horizontal and vertical competition, including RRC and EDM effects.  We estimate that without divestitures, consumers would have faced approximately \$16 million in annual harm in the Chattanooga market.  Furthermore, we show that modeling the merger as a purely horizontal upstream or downstream transaction would miss key vertical impacts.

Our paper is related to the literature on methods for measuring the welfare impacts of mergers, particularly papers about vertical mergers.  The closest merger simulation article is our previous work in \cite{ST2021}, which uses the same model we feature here but assumes that no firm is vertically integrated pre-merger.  That paper in turn takes the setup from \cite{WF1994} to characterize downstream competition, and pairs it with a Nash Bargaining model for the upstream market.  \cite{DVDS2018} and \cite{DS2019} provide additional vertical merger frameworks that make alternative assumptions about demand and pricing.  \cite{MS2013} and \cite{Rogerson2020} develop UPP-style tools for assessing the impact of vertical mergers.  However, none of these papers study complex mergers.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Theory\label{sec:theory}}
We begin by describing our basic framework, taken from \cite{ST2021}, which uses a downstream Bertrand logit model embedded in an upstream Nash Bargaining setup.  This framework allows us to study a variety of complex merger configurations.

@@

Together the series of downstream and upstream first order conditions determine market equilibrium.  This model can be solved and calibrated as described in \cite{ST2021}.  In our baseline configuration, we assume that wholesale and retail marginal costs are constant.  We allow for linear increasing marginal costs as a robustness check that appears in \ref{app:mc}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Mergers}
Here we describe the manner in which complex mergers are modeled in this framework.  We begin by describing vertical mergers, as all cases we examine have some vertical aspects.  Suppose that retailer $r$ and wholesaler $w$ were to merge.  Then the first order condition for profit maximization for the joint firm when setting the downstream price for product $rw$ is given by
\begin{equation}
\begin{split}
&\sum_{x \in \mathbb{W}^r \setminus \{w\}} [p_{rx} - p^W_{rx} - c^R_{rx}]\frac{\partial s_{rx}}{\partial p_{rw}} + s_{rw} + \overbrace{[p_{rw} - c^W_{rw} - c^R_{rw}]\frac{\partial s_{rw}}{\partial p_{rw}}}^{\text{EDM effect}} \\
&+ \underbrace{\sum_{t \in \mathbb{R}^w \setminus \{r\}} [p^W_{tw} - c^W_{tw}] \frac{\partial s_{tw}}{\partial p_{rw}}}_{\text{upstream UPP effect}} = 0.
\end{split}
\label{eq: vmerger own ret downstream foc}
\end{equation}
The pricing problem balances two effects.  On the one hand, the term labeled ``EDM effect'' captures the impact of retailer $r$ being able to access product $w$ at marginal cost.  This force would tend to lower the resulting price.  On the other hand, the term labeled ``upstream UPP effect'' captures the incentive to raise prices for retailer $r$ in order to divert sales to wholesaler $w$.  This force would tend to raise the resulting consumer price.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Numerical Simulations \label{sec:sims}}
In this section, we use a similar simulation setup to that in \cite{ST2021}, but allow for either the merging parties or their rivals to be vertically integrated before they merge. This modification allows us to explore the welfare implications of additional types of mergers beyond purely horizontal and purely vertical combinations.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Data Generating Process}
Broadly speaking, we consider four categories of complex mergers: downstream, upstream, vertical, and integrated.  We define a downstream merger as a merger between two unintegrated retailers or between an unintegrated retailer and a vertically integrated wholesaler/retailer combination.  Similarly, upstream mergers are those between two unintegrated wholesalers or between an unintegrated wholesaler and a vertically integrated wholesaler/retailer.  We define vertical mergers as those between an unintegrated wholesaler and unintegrated retailer.  Finally, an integrated merger is between two firms that are both already vertically integrated pre-merger.

@@

This paper relaxes an assumption made in much of the merger simulation literature: that mergers occur in vertical supply chains where in the pre-merger state, no firms are vertically integrated. Removing this limitation is important as competition authorities are routinely called upon to investigate complex mergers where either the merging parties or a third party are vertically integrated. Using a series of numerical simulations and the model of \cite{ST2021}, we find that most mergers involving one or more firms that are already vertically integrated harm consumers.  The exceptions occur when wholesalers have more bargaining power compared to retailers, which are instances where the benefits from EDM would tend to be larger.  We also show that having integrated non-merging rivals limits the range of welfare outcomes and tends to make mergers somewhat more beneficial to consumers, although these firms' presence is not a panacea for consumer harm. We apply our model to the Republic/Santek merger, where we find that ignoring existing vertical relationships gives an incomplete picture of the merger's welfare impacts.

Our results suggest that mergers that have horizontal aspects remain harmful for consumers in many instances, even when these mergers also have vertical aspects.
This appears to be especially true when retailers have relatively more bargaining power, which in our model occurs when retailer margins are relatively large compared to wholesaler margins.  Antitrust agencies may have reason to be especially cautious when reviewing mergers in these situations.

However, we of course have the caveat that our results are specific to our modeling framework.  Certain alternative structures are likely to give different welfare results.  Indeed, we found in \cite{ST2021} that using a second score auction to model downstream competition, as opposed to assuming Bertrand pricing, could meaningfully impact the estimated consumer harm from mergers, although results were similar between these models for vertical mergers.  This paper is an attempt to provide some information about possible welfare effects from complex mergers within a framework that is used by practitioners.  Exploring other commonly used frameworks would be a natural extension in future research.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Increasing Marginal Costs\label{app:mc}}
